Assessing the potential of Jacobian-free Newton-Krylov methods for cell-centred finite volume solid mechanics

### 2025-01-30 article.tex
cantilever - update plots
Updates

@@

The current work considers three forms of the problem:
\begin{enumerate}[label=\roman*.]
	\item Small strain, linear elastic \cite{Zienkiewicz2000, Simplas}: $E=70$ MPa, $\nu=1/3$, and $\tau = 6.25$ kPa.
	\item Finite strain, neo-Hookean hyperelastic \cite{Pelteret2018}: $E=1.0985$ MPa, $\nu=0.3$, and $\tau = 62.5$ kPa.
	\item Finite strain, neo-Hookean hyperelastoplastic \citep{Simo1992, Simplas, Cesar2001}: $E=206.9$ MPa, $\nu=0.29$, $\sigma_y = 0.45 + 0.12924\bar{\varepsilon}_p + (0.715 - 0.45)(1- e^{-16.93\bar{\varepsilon}_p})$ MPa, and $\tau = 312.5$ kPa.
\end{enumerate}
where $E$ is the Young's modulus, $\nu$ is the Poisson's ratio, $\sigma_y$ is the yield strength, $\bar{\varepsilon}_p$ is the equivalent plastic strain, and $\tau$ is the prescribed shear traction.
%The Young's modulus $E = 206.9$ MPa and Poisson's ratio $\nu=0.29$, with the yield stress $\sigma_y$ given as \citep{Simo1992}
%\begin{eqnarray}
%	\sigma_y = \sigma_Y + H\bar{\varepsilon}_p + (\sigma_{\infty} - \sigma_Y)(1- e^{-\delta\bar{\varepsilon}_p})
%\end{eqnarray}
%The plastic yielding parameters are $\sigma_Y = 0.45$ MPa, $\sigma_{\infty} = 0.715$ MPa, $\delta = 16.93$, and $H = 0.12924$ MPa, where the hardening variable $\bar{\varepsilon}_p$ corresponds to equivalent plastic strain.
%\noindent
%\textbf{Sources for case i:} \cite{Zienkiewicz2000}\\
%O.C. Zienkiewicz, R.L. Taylor. The finite element method. Butterworth Heinemann, 2000.\\
%https://www.simplassoftware.com/benchmarks.html\#biblio-58\\
%https://github.com/spolanski/CoFEA/tree/master/benchmarks/02-cooks-membrane\\
%\textbf{Sources for case ii:} \cite{dealII95}\\
%https://www.dealii.org/current/doxygen/deal.II/code\_gallery\_Quasi\_static\_Finite\_strain\_Compressible\_Elasticity.html\\
%\textbf{Sources for case iii:} \citep{Simo1992}
Eight successively refined quadrilateral meshes are considered, where the cell counts are 9, 36, 144 (Figure \ref{fig:cooks_membrane}(b)), 576, $2\,304$, $9\,216$, $36\,864$, and $147\,456$.
%mesh.1 - 3x3 - 9 CVs\\
%mesh.2 - 6x6 - 36 CVs\\
%mesh.3 - 12x12 - 144 CVs\\
%mesh.4 - 24x24 - 576 CVs\\
%mesh.5 - 48-48 - 2304 CVs\\
%mesh.6 - 96x96 - 9216 CVs\\
%mesh.7 - 192x192 - 36864 CVs
%mesh.8 - 384x384 - 147456 CVs
The problem is solved quasi-statically, and there are no body forces.
The linear elastic and hyperelastic cases use one loading increment, whereas the hyperelastoplastic case uses 30 equally-sized loading increments.

@@

Figure \ref{fig:cooksMembrane_disp} compares the predicted vertical displacement at the reference point as a function of the average cell widths with results from the literature.
The reference point is taken as the top right point -- $(48,60)$ mm -- in the elastic and hyperelastoplastic cases, while it is taken as the midway point of the loading surface -- $(48,52)$ mm -- in the hyperelastic case.
%\hl{Comment on results -> good/bad/etc}
%\hl{Comment on JFNK not working for plasticty}\\
%\hl{Philip, I'll leave this to you to comment on, as you'll be able to explain it better.}
\begin{figure}[htbp]
   \centering
   	\subfigure[Linear elastic case]
	{
		\includegraphics[width=0.45\textwidth]{figures/cooksMembrane-hookeanTipDispConvergence}  
	}
	\subfigure[Hyperelastic case]
	{
		\includegraphics[width=0.45\textwidth]{figures/cooksMembrane-neoHookeanTipDispConvergence}  
	}
	\subfigure[Hyperelastoplastic case]
	{
		\includegraphics[width=0.45\textwidth]{figures/cooksMembrane-neoHookeanPlasticTipDispConvergence}  
	}
%	\subfigure[Equivalent stress distribution for the hyperelastic case (case ii) using the mesh with $36\,842$ cells]
%	{
%		\includegraphics[height=0.5\textwidth]{figures/cooksMembrane-sigmaEq.pdf}  
%	}
%	\subfigure[Convergence of tip (48,60) vertical displacement, normalised using the displacement value from the finest mesh (given in parentheses)]
%	{
%		\includegraphics[height=0.35\textwidth]{figures/cooksMembrane-tipDispConvergence.pdf} 
%	}
   \caption{Cook's membrane vertical displacement predictions at the reference point as a function of the average cell width. Comparisons are given with the results from the literature \cite{Zienkiewicz2000, Pelteret2018, Simo1992, Simplas}. \hl{Ref NOT tip}}
   \label{fig:cooksMembrane_disp}
\end{figure}

@@

\paragraph{Case 6: Vibration of a 3-D Cantilevered Beam}
This 3-D, dynamic, finite strain case geometry consists of a $2 \times 0.2 \times 0.2$ m cuboid column and was proposed in its initial form by \citet{Tukovic2007}.
%(Figure \ref{fig:dynamic_cantilever})
%A sudden, constant traction $\bb{T} = \left(\sfrac{|\bb{T}|}{\sqrt{2}}\right) \left(1, 1, 0 \right)$ Pa is applied to the upper surface, where the magnitude $|\bb{T}|$ is defined in terms of a dimensionless load factor $\mathcal{F}$ as
%\begin{eqnarray}
%	|\bb{T}| = \frac{\mathcal{F} E I}{A L^2} \text{ Pa}
%\end{eqnarray}
A sudden, constant traction $\bb{T} = \left(80, 80, 0 \right)$ kPa is applied to the upper surface.
%A sudden, constant traction $\bb{T} = \left(\sfrac{|\bb{T}|}{\sqrt{2}}\right) \left(1, 1, 0 \right)$ Pa is applied to the upper surface, where the magnitude $|\bb{T}|$ is defined in terms of a dimensionless load factor $\mathcal{F}$ as
%In the current study, $\mathcal{F} = 1$.
A neo-Hookean hyperelastic material is assumed with $E = 15.293$ MPa, $\nu = 0.3$ and density $\rho = 1000$ kg m$^{-3}$, while the area $A = 0.04$ m$^2$ and the second moment of area $I = \nicefrac{1}{7\,500}$ m$^4$. %0.0001333
The geometric and material parameters were chosen for the first natural frequency to be 1 Hz in the small strain limit.
The magnitude of the applied traction is chosen here to test the robustness of the solution procedures when large rotations and strains are present.
Four succesively refined hexahedral meshes are generated using the OpenFOAM \texttt{blockMesh} utility, with cell counts of 270, $2\,160$, $17\,280$ and $138\,240$.
%A fixed time step size of 0.001617274174 s is chosen, corresponding to a Courant number of six on the mesh with $2\,160$ cells.
%A fixed time step size of 1 ms is chosen.
The time step size is 1 ms, and the total period is 1 s.
% corresponding to 10 expected oscillation periods.
%, corresponding to a Courant number of six on the mesh with $2\,160$ cells.

The deformed configuration of the beam at six time steps is shown in Figure \ref{fig:dynamic_cantilever}(a) for the mesh with $2\,160$ cells, where the upper surface is seen to drop below the horizontal from approximately t = 0.2 s to 0.32 s.
The corresponding displacement magnitude of the centre of the upper surface of the beam vs time is shown in Figure \ref{fig:dynamic_cantilever}(b), where the predictions are seen to converge to the results from finite element software Abaqus (element code C3D8) using the $138\,240$ mesh.
%Although the current case does not have a reference solution, it has been chosen to test the robustness of the solution procedures when large rotations and strains are present.
\begin{figure}[htbp]
   \centering
%   \includegraphics[width=0.5\textwidth]{figures/cantilever} 
	\subfigure[Displacement magnitude for t = $\left\{0, \, 0.1, \, 0.2\right\}$ s (translucent) and t = $\left\{0.3, \, 0.4, \, 0.5 \right\}$ s (opaque) for the mesh with $138\,240$ cells]
	{
	   \includegraphics[width=0.5\textwidth]{figures/cantilever_deformed_v2}
   	} \quad
	\subfigure[Displacement magnitude of the centre of the upper surface of the beam vs time]
	{
	   \includegraphics[width=0.6\textwidth]{figures/cantilever_disp} 
   	   %\includegraphics[height=0.45\textwidth]{figures/cantilever_sigmaEq}  
   	}
   \caption{Deflection of a 3-D Cantilevered Beam}
   \label{fig:dynamic_cantilever}
\end{figure}
%\hl{It would be good to add a reference results, e.g. from FE}

@@

Figure \ref{fig:hronTurek-results} shows the velocity field in the fluid region and displacement magnitude in the solid region at $t = 4.42$ s, corresponding to a peak in the vertical displacement of the plate free-end oscillation.
The fluid is seen to accelerate as it passes the cylinder, with regular vortices being thrown off the plate, causing it to oscillate.
\begin{figure}[htbp]
   \centering
	   \includegraphics[width=\textwidth]{figures/hronTurek-results} 
%	\subfigure[Velocity field in the fluid region and displacement magnitude in the solid region at $t = 4.42$ s]
%	{
%	   \includegraphics[width=0.8\textwidth]{figures/hronTurek-results} 
%   	}
%	\subfigure[Vertical displacement of the end of the plate vs time]
%	{
%   		\includegraphics[width=0.3\textwidth]{figures/placeholder}  
%   	}
   \caption{Elastic plate behind a rigid cylinder case results at $t = 20$ s for the mesh with $20\,032$ cells in the fluid region and $2\,496$ in the solid region. The velocity magnitude is shown in the fluid region while the displacement magnitude is shown in the solid region. \hl{update: harmonic top/bot}}
   \label{fig:hronTurek-results}
\end{figure}
The predicted mean, amplitude and frequency of the vertical and horizontal displacement of the end of the plate are compared with the results from \citet{Turek2006} in Table \ref{tab:hronTurekDisp}.
As directed in \citet{Turek2006}, the maximum and minimum values for the last period are used to calculate the mean and amplitude:
\begin{eqnarray}
	\text{mean} = \frac{1}{2}(\text{max} + \text{min}), \quad\quad
	\text{amplitude} = \frac{1}{2}(\text{max} - \text{min})
\end{eqnarray}
while the frequency is calculated as the inverse of the period.
\begin{table}[htb]
	\centering
		\begin{tabular}{lll}
			\hline
			Mesh (number of fluid and solid cells) & $u_x$ (in mm) & $u_y$ (in mm) \\
			\hline
			1 ($1\,252$ + 156)  & $ -0.1 \pm 0.001 \,[ 47.6 ]$ & $1.58 \pm 0.002 \,[ 5.88 ]$ \\
			2 ($5\,008$ + 624)  & $-2.16 \pm 2.00 \,[ 10.2 ]$ & $1.54 \pm 29.56 \,[ 5.6 ]$ \\
			3 ($20\,032$ + $2\,496$)  & $-2.51 \pm 2.37 \,[12.1]$ & $0.87 \pm 33.08 \,[ 5.6 ]$ \\
			\hline
			\citet{Turek2006} & $-2.69 \pm 2.53\,[10.9]$ & $1.48 \pm 34.38\,[5.3]$ \\
			\hline
		\end{tabular}
	\caption{Average predicted displacements (mean $\pm$ amplitude $[$frequency$]$) of the free end of the plate. Note that large-scale plate oscillations in mesh 1 (coarsest mesh) die out within the first 10 s.}
%	 \hl{Hron-Turek say to use the last cycle, but surely taking the average of cycles would be better for benchmarking...} \hl{e.g. the average of the mesh 3 cycle is slightly lower than the rest} \hl{we could also report the average}
		\label{tab:hronTurekDisp}
\end{table}

>>>>

%%--------------------------------------------------------------------------------------------------------------------%%
\subsection{Checking Convergence}
%%--------------------------------------------------------------------------------------------------------------------%%
\hl{sTol vs fTol vs aTol}
\hl{sTol: not equivalent in JFNK and Picard}

@@

The current work considers three forms of the problem:
\begin{enumerate}[label=\roman*.]
	\item Small strain, linear elastic \cite{Zienkiewicz2000, Simplas}: $E=70$ MPa, $\nu=1/3$, and $\tau = 6.25$ kPa.
	\item Finite strain, neo-Hookean hyperelastic \cite{Pelteret2018}: $E=1.0985$ MPa, $\nu=0.3$, and $\tau = 62.5$ kPa.
	\item Finite strain, neo-Hookean hyperelastoplastic \citep{Simo1992, Simplas, Cesar2001}: $E=206.9$ MPa, $\nu=0.29$, $\sigma_y = 0.45 + 0.12924\bar{\varepsilon}_p + (0.715 - 0.45)(1- e^{-16.93\bar{\varepsilon}_p})$ MPa, and $\tau = 312.5$ kPa.
\end{enumerate}
where $E$ is the Young's modulus, $\nu$ is the Poisson's ratio, $\sigma_y$ is the yield strength, $\bar{\varepsilon}_p$ is the equivalent plastic strain, and $\tau$ is the prescribed shear traction.
%The Young's modulus $E = 206.9$ MPa and Poisson's ratio $\nu=0.29$, with the yield stress $\sigma_y$ given as \citep{Simo1992}
%\begin{eqnarray}
%	\sigma_y = \sigma_Y + H\bar{\varepsilon}_p + (\sigma_{\infty} - \sigma_Y)(1- e^{-\delta\bar{\varepsilon}_p})
%\end{eqnarray}
%The plastic yielding parameters are $\sigma_Y = 0.45$ MPa, $\sigma_{\infty} = 0.715$ MPa, $\delta = 16.93$, and $H = 0.12924$ MPa, where the hardening variable $\bar{\varepsilon}_p$ corresponds to equivalent plastic strain.
%\noindent
%\textbf{Sources for case i:} \cite{Zienkiewicz2000}\\
%O.C. Zienkiewicz, R.L. Taylor. The finite element method. Butterworth Heinemann, 2000.\\
%https://www.simplassoftware.com/benchmarks.html\#biblio-58\\
%https://github.com/spolanski/CoFEA/tree/master/benchmarks/02-cooks-membrane\\
%\textbf{Sources for case ii:} \cite{dealII95}\\
%https://www.dealii.org/current/doxygen/deal.II/code\_gallery\_Quasi\_static\_Finite\_strain\_Compressible\_Elasticity.html\\
%\textbf{Sources for case iii:} \citep{Simo1992}
Eight successively refined quadrilateral meshes are considered, where the cell counts are 9, 36, 144 (Figure \ref{fig:cooks_membrane}(b)), 576, $2\,304$, $9\,216$, $36\,864$, and $147\,456$.
%mesh.1 - 3x3 - 9 CVs\\
%mesh.2 - 6x6 - 36 CVs\\
%mesh.3 - 12x12 - 144 CVs\\
%mesh.4 - 24x24 - 576 CVs\\
%mesh.5 - 48-48 - 2304 CVs\\
%mesh.6 - 96x96 - 9216 CVs\\
%mesh.7 - 192x192 - 36864 CVs
%mesh.8 - 384x384 - 147456 CVs
The problem is solved quasi-statically, and there are no body forces.
The linear elastic uses one loading increment, whereas the hyperelastic and hyperelastoplastic cases use 30 equally-sized loading increments.

@@

Figure \ref{fig:cooksMembrane_disp} compares the predicted vertical displacement at the reference point as a function of the average cell widths with results from the literature.
The reference point is taken as the top right point -- $(48,60)$ mm -- in the elastic and hyperelastoplastic cases, while it is taken as the midway point of the loading surface -- $(48,52)$ mm -- in the hyperelastic case.
%\hl{Comment on results -> good/bad/etc}
%\hl{Comment on JFNK not working for plasticty}\\
%\hl{Philip, I'll leave this to you to comment on, as you'll be able to explain it better.}
\begin{figure}[htbp]
   \centering
   	\subfigure[Linear elastic case]
	{
		\includegraphics[height=0.3\textwidth]{figures/cooksMembrane-hookeanTipDispConvergence}  
	}
	\subfigure[Hyperelastic case]
	{
		\includegraphics[height=0.3\textwidth]{figures/cooksMembrane-neoHookeanTipDispConvergence}  
	}
	\subfigure[Hyperelastoplastic case]
	{
		\includegraphics[height=0.3\textwidth]{figures/cooksMembrane-neoHookeanPlasticTipDispConvergence}  
	}
%	\subfigure[Equivalent stress distribution for the hyperelastic case (case ii) using the mesh with $36\,842$ cells]
%	{
%		\includegraphics[height=0.5\textwidth]{figures/cooksMembrane-sigmaEq.pdf}  
%	}
%	\subfigure[Convergence of tip (48,60) vertical displacement, normalised using the displacement value from the finest mesh (given in parentheses)]
%	{
%		\includegraphics[height=0.35\textwidth]{figures/cooksMembrane-tipDispConvergence.pdf} 
%	}
   \caption{Cook's membrane vertical displacement predictions at the reference point as a function of the average cell width. Comparisons are given with the results from the literature \cite{Zienkiewicz2000, Pelteret2018, Simo1992, Simplas}}
   \label{fig:cooksMembrane_disp}
\end{figure}

@@

\paragraph{Case 6: Vibration of a 3-D Cantilevered Beam}
This 3-D, dynamic, finite strain case geometry consists of a $2 \times 0.2 \times 0.2$ m cuboid column and was proposed in its initial form by \citet{Tukovic2007}.
%(Figure \ref{fig:dynamic_cantilever})
%A sudden, constant traction $\bb{T} = \left(\sfrac{|\bb{T}|}{\sqrt{2}}\right) \left(1, 1, 0 \right)$ Pa is applied to the upper surface, where the magnitude $|\bb{T}|$ is defined in terms of a dimensionless load factor $\mathcal{F}$ as
%\begin{eqnarray}
%	|\bb{T}| = \frac{\mathcal{F} E I}{A L^2} \text{ Pa}
%\end{eqnarray}
A sudden, constant traction $\bb{T} = \left(50, 50, 0 \right)$ kPa is applied to the upper surface.
%A sudden, constant traction $\bb{T} = \left(\sfrac{|\bb{T}|}{\sqrt{2}}\right) \left(1, 1, 0 \right)$ Pa is applied to the upper surface, where the magnitude $|\bb{T}|$ is defined in terms of a dimensionless load factor $\mathcal{F}$ as
%In the current study, $\mathcal{F} = 1$.
A neo-Hookean hyperelastic material is assumed with $E = 15.293$ MPa, $\nu = 0.3$ and density $\rho = 1000$ kg m$^{-3}$, while the area $A = 0.04$ m$^2$ and the second moment of area $I = \nicefrac{1}{7\,500}$ m$^4$. %0.0001333
The geometric and material parameters were chosen for the first natural frequency to be 1 Hz in the small strain limit.
The magnitude of the applied traction is chosen here to test the robustness of the solution procedures when large rotations and strains are present.
Four succesively refined hexahedral meshes are generated using the OpenFOAM \texttt{blockMesh} utility, with cell counts of 270, $2\,160$, $17\,280$ and $138\,240$.
%A fixed time step size of 0.001617274174 s is chosen, corresponding to a Courant number of six on the mesh with $2\,160$ cells.
%A fixed time step size of 1 ms is chosen.
The time step size is 1 ms, and the total period is 1 s.
% corresponding to 10 expected oscillation periods.
%, corresponding to a Courant number of six on the mesh with $2\,160$ cells.

The deformed configuration of the beam at six time steps is shown in Figure \ref{fig:dynamic_cantilever}(a) for the mesh with $2\,160$ cells, where the upper surface is seen to drop below the horizontal from approximately t = 0.2 s to 0.32 s.
The corresponding displacement magnitude of the centre of the upper surface of the beam vs time is shown in Figure \ref{fig:dynamic_cantilever}(b), where the predictions are seen to converge to the results from finite element software Abaqus (element code C3D8) using the $138\,240$ mesh.
%Although the current case does not have a reference solution, it has been chosen to test the robustness of the solution procedures when large rotations and strains are present.
\begin{figure}[htbp]
   \centering
%   \includegraphics[width=0.5\textwidth]{figures/cantilever} 
	\subfigure[Displacement magnitude for t = $\left\{0, \, 0.1, \, 0.2\right\}$ s (translucent) and t = $\left\{0.3, \, 0.4, \, 0.5 \right\}$ s (opaque) for the mesh with $138\,240$ cells]
	{
	   \includegraphics[width=0.5\textwidth]{figures/cantilever_deformed_50}
   	} \quad
	\subfigure[Displacement magnitude of the centre of the upper surface of the beam vs time]
	{
	   \includegraphics[width=0.6\textwidth]{figures/cantilever_disp} 
   	   %\includegraphics[height=0.45\textwidth]{figures/cantilever_sigmaEq}  
   	}
   \caption{Deflection of a 3-D Cantilevered Beam}
   \label{fig:dynamic_cantilever}
\end{figure}
%\hl{It would be good to add a reference results, e.g. from FE}

@@

Figure \ref{fig:hronTurek-results} shows the velocity field in the fluid region and displacement magnitude in the solid region at $t = 4.42$ s, corresponding to a peak in the vertical displacement of the plate free-end oscillation.
The fluid is seen to accelerate as it passes the cylinder, with regular vortices being thrown off the plate, causing it to oscillate.
\begin{figure}[htbp]
   \centering
	   \includegraphics[width=\textwidth]{figures/hronTurek-results} 
%	\subfigure[Velocity field in the fluid region and displacement magnitude in the solid region at $t = 4.42$ s]
%	{
%	   \includegraphics[width=0.8\textwidth]{figures/hronTurek-results} 
%   	}
%	\subfigure[Vertical displacement of the end of the plate vs time]
%	{
%   		\includegraphics[width=0.3\textwidth]{figures/placeholder}  
%   	}
   \caption{Elastic plate behind a rigid cylinder case results at $t = 20$ s for the mesh with $20\,032$ cells in the fluid region and $2\,496$ in the solid region. The velocity magnitude is shown in the fluid region while the displacement magnitude is shown in the solid region. \hl{update: harmonic top/bot}}
   \label{fig:hronTurek-results}
\end{figure}
The predicted mean, amplitude and frequency of the vertical and horizontal displacement of the end of the plate are compared with the results from \citet{Turek2006} in Table \ref{tab:hronTurekDisp}.
As directed in \citet{Turek2006}, the maximum and minimum values for the last period are used to calculate the mean and amplitude:
\begin{eqnarray}
	\text{mean} = \frac{1}{2}(\text{max} + \text{min}), \quad\quad
	\text{amplitude} = \frac{1}{2}(\text{max} - \text{min})
\end{eqnarray}
while the frequency is calculated as the inverse of the period.
\begin{table}[htb]
	\centering
		\begin{tabular}{lll}
			\hline
			Mesh (number of fluid and solid cells) & $u_x$ (in mm) & $u_y$ (in mm) \\
			\hline
			\input{tableHronTurekDisp.tex}
%			1 ($1\,252$ + 156)  & $ -0.1 \pm 0.001 \,[ 47.6 ]$ & $1.58 \pm 0.002 \,[ 5.88 ]$ \\
%			2 ($5\,008$ + 624)  & $-2.16 \pm 2.00 \,[ 10.2 ]$ & $1.54 \pm 29.56 \,[ 5.6 ]$ \\
%			3 ($20\,032$ + $2\,496$)  & $-2.51 \pm 2.37 \,[12.1]$ & $0.87 \pm 33.08 \,[ 5.6 ]$ \\
%			\hline
			\citet{Turek2006} & $-2.69 \pm 2.53\,[10.9]$ & $1.48 \pm 34.38\,[5.3]$ \\
			\hline
		\end{tabular}
	\caption{Predicted oscillations (mean $\pm$ amplitude $[$frequency$]$) of the free end of the plate. Note that large-scale plate oscillations in mesh 1 (coarsest mesh) die out within the first 10 s.}
%	 \hl{Hron-Turek say to use the last cycle, but surely taking the average of cycles would be better for benchmarking...} \hl{e.g. the average of the mesh 3 cycle is slightly lower than the rest} \hl{we could also report the average}
		\label{tab:hronTurekDisp}
\end{table}
